Solutions for decentralized AI\\

### 2023-06-19 paper.tex
Paper 100%

\section{Introduction}
In recent years, AI models have become a relevant technology: from chatbots to image generators, and many other impressive applications to come.

These technologies are usually developed by companies that take on the role of developing Deep Learning models, gathering the training data, and training the models with their own computing resources. In the future, it will be no surprise to also see governments taking a leading role in the development and use of Deep Learning technologies.

With these two scenarios in mind, it is natural to fear the power that companies and governments can accumulate against citizens by owning these models. Many philosophers, journalists, writers, anthropologists, historians, lawyers, and economists have already pointed out how the tech industry is becoming highly centralized and monopolistic, and how this trend can lead to more economic inequality and authoritarian governments \cite{b1} \cite{b2} \cite{b3} \cite{b4}.

Luckily, at the same time, another technology that seeks decentralization has been emerging: Blockchain. This technology started as a decentralized network of computers that maintained an append-only immutable ledger of transactions \cite{b6} but has developed into a wider concept allowing even other kinds of decentralized applications.

@@

\section{Our contribution}
First, we propose a consensus protocol that uses Deep Learning training as Proof of Useful Work. We provide an impossibility result that shows why such a protocol (with the assumptions and properties we desire) would not be feasible in practice.

We provide an alternative solution using smart contracts to train models and provide token-based incentives to the trainers and the data providers. We also show empirically that this alternative is very costly in terms of gas.

Finally, we pose Layer-2 applications as a more realistic alternative to the decentralization of AI models.

\section{Background, terminology and definitions}
\subsection{Blockchain background}
Blockchains are a solution to the State Machine Replication (SMR) problem. In Blockchain networks, different nodes try to reach a consensus as to what is the true state of the system (whether it is a set of transactions, balances, variable values, or, related to this paper, the most optimal parameters of a Deep Learning model) even in the presence of Byzantine nodes.

Consensus protocols in the form of Proof of Work revolve around the idea of miners competing to solve a problem that is hard to solve but easy to verify. Bitcoin's PoW uses the problem of breaking SHA-256 hashes as the problem that miners have to solve in order to be the leaders of a round.

@@

Another concept in the world of Blockchains is Smart Contracts. Smart Contracts are code deployed in the Blockchain that can be executed by anyone. Thanks to them being in the Blockchain, no one can restrict access to their functions and no one can manipulate its output.

@@

We use the term Deep Learning model to refer to a certain Deep Learning architecture with some parameters $W$ that have been trained with training data or are still to be trained. Model parameters are initialized randomly and trained iteratively using some optimization algorithm like stochastic gradient descent.

@@

The training of the model depends on the choice of some hyperparameters. These hyperparameters include the learning rate $\alpha$, the optimization algorithm, or the number and size of layers in the neural network. These can influence both the results and the duration of the training.

@@

\section{Embedding decentralized AI in the consensus protocol}
\subsection{Previous work}
There have been many attempts to develop Proof of Useful Work protocols that use Deep Learning training as the hard problem to be solved by miners.

@@

Except for Proof-of-Deep-Learning (2019), where they assumed a single supplier that should be honest, all the other suggested protocols require problem suppliers to provide a reward for the winning miner. The authors of CrowdMine even propose burning part of the rewards instead of giving them in full to the winner.

@@

The protocol runs for sequential rounds at the end of which all honest nodes should agree on one block to be appended to their local histories. Each round has three phases:
\begin{enumerate}
    \item ``Training phase'', in which:
    \begin{enumerate}
        \item the submitter broadcasts a training dataset $\{\mathcal X_\text{train},\mathcal Y_\text{train}\}$ to all miners;
        \item miners train their models $\{f,W\}^{(i)}$ using the training data, where $i$ denotes the $i$'th miner;
        \item miners have to commit to their final model $\{f,W\}^{(i)}$ by sending a hash $h^{(i)}$ of it to the submitter before a timeout.
    \end{enumerate}
    \item ``Evaluation phase'', in which:
    \begin{enumerate}
        \item the submitter broadcasts the test input $\mathcal X_\text{test}$ to all miners;
        \item miners compute their estimates for the test output using their model $f^{(i)}(x_{\text{test}}|W^{(i)})=\hat y_{\text{test}}^{(i)}$;
        \item miners send their estimates $\hat{\mathcal Y}_{\text{test}}^{(i)}$ to the submitter (this serves as a Zero-Knowledge proof for the submitter that the miner has a way to predict $y$ from $x$ with some deterministic accuracy);
        \item the submitter computes the accuracy $a^{(i)}$ of the model of each miner $i$ by comparing their estimates $\hat{\mathcal Y}_{\text{test}}^{(i)}$ with the real test dataset $\mathcal Y_{\text{test}}$.
    \end{enumerate}
    \item ``Awarding phase'', in which:
    \begin{enumerate}
        \item the submitter ranks miners by their accuracy;
        \item the submitter sends a signed message to the miner with the most accuracy (denoted $w$) certifying it as the winner of the round;
        \item the submitter sends the hash $h^{(w)}$ of the model and the estimate $\hat{\mathcal Y}_{\text{test}}^{(w)}$ of the winner to all the miners;
        \item the winner is then allowed to build a block in which he will include the signed certificate of the submitter and the model $\{f,W\}^{(w)}$;
        \item the block is broadcasted to all other miners;
        \item miners verify the block by checking that:
        \begin{enumerate}
            \item it includes the certificate of the submitter of the round,
            \item the hash of the published model $\{f,W\}^{(w)}$ corresponds with the hash $h^{(w)}$ the winner committed to during the training phase,
            \item and that the published model $\{f,W\}^{(w)}$ indeed maps $\mathcal X_\text{test}$ to the estimate $\hat{\mathcal Y}_{\text{test}}^{(w)}$ to which the winner committed to during the evaluation phase.
        \end{enumerate}
        If the published model does not correspond with the hash or is not the model that the winner used to compute $\hat{\mathcal Y}_{\text{test}}^{(w)}$, honest nodes will not add the block to their local histories and will notify the misconduct to the submitter that will choose the next model with best accuracy and will start again from step 3b.

@@

@@

If all nodes are honest, we see that the share of rewards after ... rounds is equal.

\begin{figure}[htbp]
\centerline{\includegraphics{fig1.png}}
\caption{Example of a figure caption.}
\label{fig}
\end{figure}

With one Byzantine node submitting pre-solved problems, we see that the dishonest player earns more rewards than the rest.

\begin{figure}[htbp]
\centerline{\includegraphics{fig1.png}}
\caption{Example of a figure caption.}
\label{fig}
\end{figure}

So, from the proof sketch and the experiment, we conclude that without rewards and/or fees provided by the submitters, miners can manipulate their probability of mining a block and thus win the reward.

@@

\begin{itemize}
\item Galaxy Learning (2019)\cite{} is the first attempt known to us to decentralize data storage and AI model training at the same time. They also propose incentives to data providers as a way for individuals to leverage their power against companies or institutions that regularly make use of their data. For the training mechanism, they base it on the concept of federated learning.
\item The Microsoft's SUM project (2019)\cite{} also proposes a smart-contract-based framework for sharing and improving machine learning models. They define a way to incentivize users to provide more training data and punish wrong data submissions.
\item In Incentivized Decentralized Machine Learning (2023)\cite{} and BRAIN (2023)\cite{}, the training is done by trainers off-chain and they then propose updates using a commit-reveal mechanism. The smart contract then scores and rewards the updates that are most optimal.
\end{itemize}

@@

We wanted to test how expensive it would be to train a simple model like a linear regression model in a smart contract.

We store the training data in the smart contract in two arrays $x$ and $y$, and implement a function ``train()'' that runs stochastic gradient descent to update the values of the parameters $w_0$ and $w_1$.

The smart contract code can be found in GitHub: \url{https://github.com/SergiHernandez/DecentralizedAI}

@@

\begin{table}[htbp]
\caption{Table Type Styles}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
\cline{2-4} 
\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
\hline
copy& More table copy$^{\mathrm{a}}$& &  \\
\hline
\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
\end{tabular}
\label{tab1}
\end{center}
\end{table}

@@

We have seen that it is impossible to have a decentralized deep learning system with the properties we suggest implemented in the consensus protocol of the blockchain. We have also seen that the gas costs of implementing a deep learning training algorithm in a smart contract would be incredibly economically high.

Therefore, we see that the only alternatives are:
\begin{itemize}
\item Relaxing the properties and assumptions so that we are able to implement decentralized AI in the consensus protocol (like in BlockML or PoLe).
\item Keeping the properties and assumptions and use a smart contract solution but move to a Layer 2 setting that allows greater scalability and reduction of gas costs.
\end{itemize}

@@

>>>>

\section{Introduction}
In recent years, the rapid advancements in Deep Learning have demonstrated the potential of AI models across various applications: from chatbots to image generators and beyond.

However, these technologies are usually owned and developed by companies that take on the role of designing the models and the algorithms, gathering data, and training the models with their own computing resources. In the future, it will be no surprise to also see governments taking a leading role in the development and use of Deep Learning technologies.

With these two scenarios in mind, it is natural to fear the power that companies and governments can accumulate against citizens by owning these models. Many experts from diverse fields such as philosophy, journalism, anthropology, law, and economics have already highlighted the risks associated with a highly centralized and monopolistic tech industry, including increased economic inequality and the potential for more authoritarian governments  \cite{b1} \cite{b2} \cite{b3} \cite{b4}.

Fortunately, alongside these developments, another technology that seeks decentralization has been emerging: Blockchain. This technology started as a decentralized network of computers that maintained an append-only immutable ledger of transactions \cite{b6} but has developed into a wider concept allowing even other kinds of decentralized applications.

@@

\section{Our contribution}
First, we propose a consensus protocol that uses Deep Learning training as Proof of Useful Work. We present an impossibility result that demonstrates the practical challenges associated with such a protocol under our desired assumptions and properties.

We provide an alternative solution using smart contracts to train models and provide token-based incentives to the trainers and the data providers. Through empirical analysis, we demonstrate that this solution incurs significant costs in terms of gas.

Finally, we pose Layer-2 applications as a more realistic and scalable alternative to the decentralization of AI models.

By presenting these contributions, our research provides valuable insights into the challenges and potential solutions for decentralizing AI models, paving the way for future research about more democratized and inclusive AI ecosystems.

\section{Background, terminology and definitions}
\subsection{Blockchain background}
Blockchains address the State Machine Replication (SMR) problem, where multiple nodes aim to reach consensus on the true state of a system (whether it is a set of transactions, balances, variable values, or, related to this paper, the most optimal parameters of a Deep Learning model), even in the presence of Byzantine nodes.

Consensus protocols in that use Proof of Work revolve around the idea of miners competing to solve a problem that is hard to solve but easy to verify. Bitcoin's PoW uses the problem of breaking SHA-256 hashes as the problem that miners have to solve in order to be the leaders of a round.

@@

Smart Contracts are another significant concept in the world of Blockchains. Smart Contracts are code deployed in the Blockchain that can be executed by anyone. Thanks to them being in the Blockchain, no one can restrict access to their functions and no one can manipulate its output.

@@

We use the term Deep Learning model to refer to a certain Deep Learning architecture with trainable parameters $W$ that are initialized randomly and trained iteratively using some optimization algorithm like stochastic gradient descent.

@@

The training of the model depends on the choice of some hyperparameters. These hyperparameters include the learning rate $\alpha$, the optimization algorithm, or the number and size of layers in the neural network. These can influence both the results and the duration of the training process.

@@

\section{Embedding decentralized AI in the consensus protocol}
\subsection{Previous work}
Several attempts have been made to develop Proof of Useful Work protocols that leverage Deep Learning training as the computationally intensive task for miners:

@@

Except for Proof-of-Deep-Learning (2019), where they assumed a single supplier that should be honest, all the other suggested protocols require problem suppliers to provide a reward for the winning miner. The authors of CrowdMine even propose burning part of the rewards instead of giving them in full to the winner to compensate from MEV incentives.

@@

The protocol runs for sequential rounds at the end of which all honest nodes should agree on one block to be appended to their local histories. Each round has three phases:
\begin{enumerate}
    \item ``Training phase'', in which:
    \begin{enumerate}
        \item the submitter broadcasts a training dataset $\{\mathcal X_\text{train},\mathcal Y_\text{train}\}$ to all miners;
        \item miners train their models $\{f,W\}^{(i)}$ using the training data, where $i$ denotes the $i$'th miner;
        \item miners have to commit to their final model $\{f,W\}^{(i)}$ by sending a hash $h^{(i)}$ of it to the submitter before a timeout.
    \end{enumerate}
    \item ``Evaluation phase'', in which:
    \begin{enumerate}
        \item after the timeout, the submitter broadcasts the test input $\mathcal X_\text{test}$ to all miners;
        \item miners compute their estimates for the test output using their model $f^{(i)}(x_{\text{test}}|W^{(i)})=\hat y_{\text{test}}^{(i)}$;
        \item miners send their estimates $\hat{\mathcal Y}_{\text{test}}^{(i)}$ to the submitter (this serves as a Zero-Knowledge proof for the submitter that the miner has a way to predict $y$ from $x$ with some deterministic accuracy);
        \item the submitter computes the accuracy $a^{(i)}$ of the model of each miner $i$ by comparing their estimates $\hat{\mathcal Y}_{\text{test}}^{(i)}$ with the real test dataset $\mathcal Y_{\text{test}}$.
    \end{enumerate}
    \item ``Awarding phase'', in which:
    \begin{enumerate}
        \item the submitter ranks miners by their accuracy;
        \item the submitter sends a signed message to the miner with the most accuracy (denoted $w$) certifying it as the winner of the round;
        \item the submitter sends the hash $h^{(w)}$ of the model and the estimate $\hat{\mathcal Y}_{\text{test}}^{(w)}$ of the winner to all the miners;
        \item the winner is then allowed to build a block in which he will include the signed certificate of the submitter and the model $\{f,W\}^{(w)}$;
        \item the block is broadcasted to all other miners;
        \item miners verify the block by checking that:
        \begin{enumerate}
            \item it includes the certificate of the submitter of the round,
            \item the hash of the published model $\{f,W\}^{(w)}$ corresponds with the hash $h^{(w)}$ the winner committed to during the training phase,
            \item and that the published model $\{f,W\}^{(w)}$ indeed maps $\mathcal X_\text{test}$ to the estimate $\hat{\mathcal Y}_{\text{test}}^{(w)}$ to which the winner committed to during the evaluation phase.
        \end{enumerate}
        If the published model does not correspond with the hash or is not the model that the winner used to compute $\hat{\mathcal Y}_{\text{test}}^{(w)}$, honest nodes will not add the block to their local histories and will notify the misconduct to the submitter that will choose the next model with best accuracy and will start again from step 3b.

@@

As we can see in Figure 1, if all five nodes are honest and have the same computational power, the share of rewards after 1000 rounds is equally split among them ($\approx20\%$ each).

\begin{figure}[htbp]
\centerline{\includegraphics[scale=0.50]{share-rewards-all-honest.png}}
\caption{Share of rewards after round 1000 with all nodes being honest.}
\label{fig}
\end{figure}

On the other hand, as we see in Figure 2, with one Byzantine node submitting pre-solved problems, the dishonest player earns more rewards than the rest. In particular, in the simulation, the Byzantine node submitted $50\%$ of the problems of the pool, making him earn $50\%$ of the rewards plus $10\%$ that come from splitting the rewards of the other problems with the rest of miners equally.

\begin{figure}[htbp]
\centerline{\includegraphics[scale=0.50]{share-rewards-one-dishonest.png}}
\caption{Share of rewards after round 1000 with one dishonest node.}
\label{fig}
\end{figure}

So, from the proof sketch and the experiment, we conclude that without rewards and/or fees provided by the submitters, miners can manipulate their probability of mining a block and thus win the newly-minted reward.

@@

@@

\begin{itemize}
\item Galaxy Learning (2019)\cite{b15} is the first attempt known to us to decentralize data storage and AI model training at the same time. They also propose incentives to data providers as a way for individuals to leverage their power against companies or institutions that regularly make use of their data. For the training mechanism, they suggest federated learning.
\item The Microsoft's SUM project (2019)\cite{b16} also proposes a smart-contract-based framework for sharing and improving machine learning models. They define a way to incentivize users to provide more training data and punish wrong data submissions.
\item In Incentivized Decentralized Machine Learning (2023)\cite{b17} and BRAIN (2023)\cite{b18}, the training is done by trainers off-chain and they then propose updates using a commit-reveal mechanism. The smart contract then scores and rewards the updates that are most optimal.
\end{itemize}

@@

To assess the feasibility of training a simple model like a linear regression model within a smart contract, we conducted a test.

The test involved storing the training data in the smart contract using two arrays, namely, $x$ and $y$. Additionally, we implemented a function called "train()" within the smart contract that utilizes stochastic gradient descent to update the parameter values $w_0$ and $w_1$.

We then deployed the Solidity smart contract using the Brownie framework and used the web3 library to approximate gas costs.

The Solidity smart contract code can be found in GitHub: \url{https://github.com/SergiHernandez/DecentralizedAI}

@@

Using the web3 library, we can approximate the gas costs of executing the linear regression smart contract.

\begin{table}[htbp]
\caption{Gas costs of on-chain linear regression$^{\mathrm{a}}$}
\begin{center}
\begin{tabular}{|c|c|c|c|}

\hline
\textbf{Action}&\textbf{Gas cost}&\textbf{Gwei cost}&\textbf{in USD} \\

\hline
addData()& 103,299 & 2,065,980 & $\$3.59$  \\ \hline
train()& 138,433 & 2,768,660 & $\$4.82$ \\  \hline
Full test execution& 276,305,223 & 5,526,104,460 & $\$9,626.47$\\
\hline
\multicolumn{4}{c} \text{$^{\mathrm{a}}$With 1 gas $\approx$ 20 gwei and 1 ETH $\approx$ 1,742 USD as of June 2023}
\end{tabular}
\label{tab1}
\end{center}
\end{table}

As we see in Table 1, the cost of fitting a line to just nine data points in the Ethereum chain can amount to thousands of dollars. To put this into perspective, executing the same linear regression algorithm in Python on a low-cost laptop would only take a few seconds and consume minimal energy, amounting to just a few cents.

@@

In this study, we explored the feasibility of implementing decentralized AI solutions within consensus protocols and smart contracts. We began by examining existing protocols and their limitations in achieving a fair and secure decentralized AI system. Through an impossibility proof, we demonstrated that it is impossible to prevent Byzantine nodes from manipulating their probability of winning rewards in a permissionless consensus protocol, even with a single Byzantine node. Additionally, we examined the possibility of implementing decentralized AI in smart contracts and evaluated the gas costs associated with executing a linear regression model. The gas costs on the Ethereum blockchain were found to be significantly higher compared to executing the same task on a traditional computing device.

Considering the limitations and challenges identified throughout our analysis, we conclude that achieving a fully decentralized and efficient AI system within consensus protocols and smart contracts remains a complex task. While previous works have proposed innovative solutions, such as federated learning and incentivization mechanisms, further research and development are required to address scalability, security, and cost-effectiveness concerns.

Therefore, our solutions are:
\begin{itemize}
\item Embedding decentralized AI in the consensus protocol but requiring submitters to provide rewards.
\item Using a smart contract approach but move to a Layer 2 setting that would allow greater scalability through the reduction of gas costs.
\end{itemize}

In conclusion, while the vision of decentralized AI holds great promise for democratizing access to AI technologies and enhancing privacy, achieving its full potential requires innovative solutions and a deep understanding of the unique challenges posed by decentralized systems.

@@

\bibitem{b15} Wu, C., Xiao, J., Huang, G., \& Wu, F. (2019). Galaxy Learning - A Position Paper. 
https://doi.org/10.48550/arXiv.1905.00753

\bibitem{b16} Harris, J. D., \& Waggoner, B. (2019, July). Decentralized and collaborative AI on blockchain. In 2019 IEEE international conference on blockchain (Blockchain) (pp. 368-375). IEEE. https://doi.org/10.1109/Blockchain.2019.00057.

\bibitem{b17} Yu, H., Chen, H. Y., Lee, S., Vishwanath, S., Zheng, X., \& Julien, C. (2023). idml: Incentivized decentralized machine learning. 
https://doi.org/10.48550/arXiv.2304.05354

\bibitem{b18} Park, S., Lee, J., \& Moon, S. M. (2023). A Blockchain-based Platform for Reliable Inference and Training of Large-Scale Models. 
https://doi.org/10.48550/arXiv.2305.04062.
